Strong lensing selection effects

### 2022-07-21 paper/seleff.tex
started section 4

\def\reff{R_{\mathrm{e}}}
\def\mstar{M_*}
\def\gammadm{\gamma_{\mathrm{DM}}}
\def\fdm{f_{\mathrm{DM}}}
\def\mdmfive{M_{\mathrm{DM}, 5}}

@@

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisit the \citet{MVK09} study, we expand it to the extended source case and, most importantly, we address point 2) as well: we simulate large populations of strong lenses using semi-analytic models, simulate lensing surveys with realistic assumptions for the lens probability $\pdet$ and quantify the lensing bias under various scenarios.
We explore how the results change as a function of the scatter in mass parameters at fixed light, and as a function of source type: we simulate both galaxy and quasar sources.

@@

@@

>>>>

\def\reff{R_{\mathrm{e}}}
\def\mstar{M_*}
\def\asps{\alpha_{\mathrm{sps}}}

\def\mobs{M_*^{(\mathrm{obs})}}
\def\gammadm{\gamma_{\mathrm{DM}}}
\def\fdm{f_{\mathrm{DM}}}
\def\mdmfive{M_{\mathrm{DM}, 5}}

@@

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisit the \citet{MVK09} study, we expand it to the extended source case and, most importantly, we address point 2) as well: we simulate large populations of strong lenses using empirical models, simulate lensing surveys with realistic assumptions for the lens probability $\pdet$ and quantify the lensing bias under various scenarios.
We explore how the results change as a function of the scatter in mass parameters at fixed light, and as a function of source type: we simulate both galaxy and quasar sources.

@@

In the previous section we studied how the lensing cross-section, which is closely related to the term $\pdet$ in \Eref{eq:one}, varies as a function of lens and source properties.
From here on we study the effect that those trends have on populations of lenses.
We do this by simulating populations of foreground galaxies and background sources, selecting strong lenses among them, and comparing the properties of the strong lens sample with the general galaxy population.
Our simulations are based on empirical models, in which existing observations of the baryonic component of galaxies are complemented with a set of assumptions on the mass distribution of the lenses.
In section \ref{ssec:lenses} we explain how we built our foreground galaxy sample,
while in section \ref{ssec:sources} we describe the simulation of the background source population.
In section \ref{ssec:obs} we describe how our mock observations of lenses are generated.

%As we explained in \Sref{sect:intro}, a key ingredient determining the amplitude of the strong lensing bias is the scatter in the mass properties of the foreground galaxy population.
%Although we do not have robust measurements of the intrinsic scatter in properties such as the stellar mass-to-light ratio or the dark matter halo mass, we set upper limits on their amplitude using the 
%Our simulations explore different amplitudes of the scatter. Scatter

%In section \ref{ssec:lenses} we explain how we built our foreground galaxy sample.
%In section \ref{ssec:

\subsection{Foreground galaxies}\label{ssec:lenses}

Our foreground galaxy population consists of a volume-limited sample of early-type galaxies, complete above a minimum observed stellar mass of $10^{11}M_\odot$.
We chose to focus on early-type galaxies because many strong lensing surveys have preferentially targeted this class of objects in their lens-finding phase \citep[e.g.][]{Bol++06,Gav++12,Son++18a}.
%This, in turn, was motivated by two points:
This, in turn, had a dual motivation:
first, early-type galaxies are among the most massive objects in the Universe, and therefore they are more likely to be lenses;
second, their smooth surface brightness distribution and red colour makes it easier to detect arcs from strongly lensed star forming galaxies around them.

We describe lenses with elliptical versions of the two-component model introduced in section \ref{ssec:profile}.
In the following sections we explain how their parameters are generated.

\subsubsection{Stellar mass and redshift distribution}

We generated lens galaxies over a finite redshift range, $0.1 < z < 0.9$.
We chose these lower and upper limits because the value of the critical surface mass density $\Sigma_{\mathrm{cs}}$ becomes very large outside of this range, and the fraction of galaxies that act as strong lenses drops substantially as a result.

We drew stellar masses from the stellar mass function of quiescent galaxies measured by \citet{Muz++13}.
In particular, we used the following comoving number density distribution
\begin{equation}
\Phi(\mobs) = (\ln{10})\Phi^*\left[10^{(\mobs - \mstar^*)(1+\alpha)}\right]\times \exp{\left[-10^{(\mobs -\mstar^*)}\right]}.
\end{equation}
We set $\Phi^*=1.009\times10^{-3}\,{\rm Mpc}^{-3}$, $\alpha=-0.92$ and $\log{\mstar^*}=11.21$. These are the best-fit values measured by \citet{Muz++13}.

We refer to the stellar masses drawn from this distribution as the observed stellar masses, $\mobs$, as opposed to the true stellar masses.
The observed stellar mass is meant to represent an estimate of $\mstar$ based on stellar population synthesis modelling, which is the method used by \citet{Muz++13} to measure the galaxy stellar mass function.
Stellar population synthesis measurements, however, are subject to systematic uncertainties, since they have not been calibrated on galaxies whose stellar mass is known by other means.
We quantify the discrepancy between the observed and true stellar mass by means of the stellar population synthesis mismatch parameter $\asps$, defined as follows:
\begin{equation}
\asps \equiv \frac{\mstar}{\mobs}.
\end{equation}

%We randomly assign a value of $\asps$ to each galaxy in the sample, by drawing it from the following distribution:
For each galaxy in the sample, we randomly drew a value of $\log{\asps}$ from the following distribution:
\begin{equation}
\pr(\log{\asps}) \sim \mathcal{N}(0.1, \sigma_\alpha^2),
\end{equation}
where the notation $\mathcal{N}(\mu,\sigma^2)$ indicates a Gaussian with mean $\mu$ and variance $\sigma^2$.
We set the mean of $\log{\asps}$ to $0.1$, as this is an intermediate value among estimates of $\asps$ from the literature \citep{CvD12, Cap++13, SLC15, Son++15, Son++19}.
For fhe scatter $\sigma_\alpha$ we adopted a few different values. We explain in \ref{ssub:scat} how these were chosen.

\subsubsection{Surface mass density distribution}

We described the surface mass density distribution of each galaxy as an elliptical de Vaucouleurs profile (a S\'{e}rsic profile with $n=4$).
Given the observed stellar mass of a galaxy, we assigned a half-mass radius by drawing it from the following distribution in $\log{\reff}$:
\begin{equation}\label{eq:masssize}
\pr(\log{\reff}) \sim \mathcal{N}(1.20 + 0.63(\log{\mobs} - 11.4), 0.14^2).
\end{equation}
Then, we assigned an axis ratio $q$ by drawing it from the following distribution:
\begin{equation}\label{eq:qdist}
\pr(q) \sim \,{\rm boh}.
\end{equation}
%The choice for the distributions of \Eref{eq:masssize} and \Eref{eq:qdist}
The choice for these distributions was motivated by observations of a sample of early-type galaxies. In Appendix~\ref{sect:appendixa} we explain how this sample was defined and how the coefficients of \Eref{eq:masssize} and \Eref{eq:qdist} were determined.

\subsubsection{Dark matter distribution}

\subsubsection{Intrinsic scatter}\label{ssub:scat}

\subsection{Background sources}\label{ssec:sources}

\subsection{Observations}\label{ssec:obs}

@@

\appendix
\section{Lens galaxy surface brightness distribution}\label{sect:appendixa}

To assign half-light radii and ellipticities to the simulated lenses we relied on observations of a sample of early-type galaxies selected from the Sloan Digital Sky Survey \citep[SDSS][]{Yor++00}.
This sample was selected as follows. First we defined a narrow redshift slice around $z=0.2$. Then we applied a selection in colour, by choosing objects with $g-r>1.2$, and on S\'{e}rsic index, by selecting only galaxies with $n>2.5$. We used the S\'{e}rsic fit measurements by \citet{Mee++15} for this purpose. These cuts produced a sample of $8078$ galaxies.
%We focused on the $r-$band de Vaucouleurs profile measurements obtained by \citet{Mee++15}, for these objects.
%We then took measurements of their total flux, half-light radius and axis ratio from the de Vaucouleurs profile fits to the SDSS $r$-band data obtained by \citet{Mee++15}. Using the 
%We then took measurements of the total flux, half-light radius and axis ratio obtained by \citet{Mee++15} from the de Vaucouleurs profile fits to the SDSS $r$-band data obtained by \citet{Mee++15}. Using the 
%For each object, we took the measurement of its de Vaucouleurs model-based $r-band$ total flux, half-light radius and axis ratio, as provided by \citet{Mee++15}.
We then focused on the $r-$band de Vaucouleurs model-based photometric measurements of \citet{Mee++15}.
Using the $r-$band total flux from the de Vaucouleurs model and the stellar mass-to-light ratio estimates of \citet{Men++14}, we obtained measurements of $\mobs$. Finally, we fitted the stellar mass-size relation with the following model:
\begin{equation}
\log{\reff} \sim \mathcal{N}(\mu_{R,0} + \beta_R(\log{\mobs} - 11.4), \sigma_R^2).
\end{equation}
We obtained
